---
title: "Statistical analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

This tutorial shows how to test associations between `CellTFusion` features — latent factors, TF modules or cell groups — and clinical variables. It uses the result of `CellTFusion()`:

```{r, eval = FALSE}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto

res <- CellTFusion(
  raw.counts  = raw.counts,
  normalized  = TRUE,
  cancer_type = "skcm",
  return      = TRUE
)
```

## **Latent factors and clinical traits**

`scores.stat.analysis()` tests each latent factor against a clinical variable and saves a plot for every significant association in `Results/`. `scores` can be the output of `compute.latent_factors()` (`res$Latent_spaces`), a samples x features matrix, or a list whose first element is that matrix (e.g. `res$Cell_groups`). The rows must be in the same order as the rows of `coldata`.

Available tests (`method`):

- `"wilcox"` — Wilcoxon rank-sum test (two groups).
- `"ttest"` — Student's t-test (two groups).
- `"kruskal"` — Kruskal-Wallis test with Dunn post-hoc tests (two or more groups).
- `"anova"` — one-way ANOVA with Tukey post-hoc tests (two or more groups).
- `"fisher"` — Fisher's exact test, with scores split at the median into High/Low.

```{r, eval = FALSE}
sig_factors <- scores.stat.analysis(
  scores  = res$Latent_spaces,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "wilcox",
  pval    = 0.05
)
```

The result keeps only the significant features (`NULL` if none). The individual tests (`scores.wilcox.test()`, `scores.ttest()`, `scores.kruskal.test()`, `scores.anova.test()`, `scores.fisher.test()`) can also be called directly with a list whose first element is the score matrix.

The factor scores themselves are in `res$Latent_spaces$Z`:

```{r, eval = FALSE}
factor_scores <- data.frame(res$Latent_spaces$Z)
head(factor_scores)
```

## **TF modules and clinical traits**

`compute.metadata.association()` relates TF module scores to all clinical variables at once:

- **Numeric traits** are correlated with each module (Pearson by default, `corr_method = "s"` for Spearman) and shown in a heatmap, annotated for p-values below `pval`. For traits with missing values, each p-value uses the number of samples with both values.
- **Categorical traits** are tested with one-way ANOVA and Tukey post-hoc tests when `plot_grid = TRUE`, and saved as boxplot grids.

```{r, eval = FALSE}
compute.metadata.association(
  tfs.modules = res$TF_network[[1]],
  coldata     = traitdata,
  pval        = 0.05,
  file.name   = "Tutorial",
  plot_grid   = TRUE
)
```

## **TF modules and pathways**

`compute.modules.relationship()` correlates two feature matrices over the same samples, for example TF modules and pathway activities, and saves a heatmap of the significant correlations. With `plot.grid = TRUE`, a scatter plot is also drawn for every significant pair.

```{r, eval = FALSE}
compute.modules.relationship(
  matA      = res$TF_network[[1]],
  matB      = res$Pathways_scores,
  file_name = "Pathways_vs_TF_modules",
  width     = 15
)
```

## **Survival analysis**

`compute.survival.analysis()` compares Kaplan-Meier survival curves with a log-rank test (packages `survival`, `survminer` and `gridExtra`). It works in two modes:

1. **Predefined groups** (`group_column`) — e.g. a clinical risk group or a cluster assignment in the clinical data.
2. **Feature screening** (`features`) — e.g. `res$Latent_spaces$Z`. Each feature is split into High/Low at a quantile (`thres`), and only features with a significant log-rank test (`p.value`) are returned. The rows of `features` must be the same samples, in the same order, as the rows of the clinical data.

Both modes need a time-to-event column and an event column (1 = event, 0 = censored) in the clinical data.

```{r, eval = FALSE}
# Predefined groups
surv_group <- compute.survival.analysis(
  survival.data = traitdata,
  PFS           = "PFS",
  PFS_event     = "PFS_event",
  group_column  = "Best.Confirmed.Overall.Response",
  file_name     = "Tutorial"
)

surv_group$p_value     # log-rank p-value
surv_group$median_PFS  # median survival per group
```

```{r, eval = FALSE}
# Feature screening on latent factors
surv_factors <- compute.survival.analysis(
  survival.data = traitdata,
  PFS           = "PFS",
  PFS_event     = "PFS_event",
  features      = res$Latent_spaces$Z,
  p.value       = 0.05,
  thres         = 0.5,   # median split; e.g. 0.75 for top quartile vs rest
  file_name     = "Tutorial"
)
```

For every significant result, a Kaplan-Meier plot with a risk table is saved to `Results/SurvPlot_<group-or-feature>_<file_name>.svg`.
