---
title: "Machine learning workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine learning workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

`CellTFusion` latent factors can be used as features to predict clinical outcomes with the [`pipeML`](https://github.com/VeraPancaldiLab/pipeML) package:

```{r, eval = FALSE}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/pipeML")
```

Latent factors are learned from the data (cell groups, TF modules, NMF), so they must be learned **within each cross-validation fold**: learning them on all the training samples before cross-validation would let the test samples of each fold influence the features and overestimate performance. `prepare_celltfusion_folds()` does this for `pipeML`: in each fold, `CellTFusion()` is run on the training samples only, and the test samples are projected onto the resulting latent factors with `project_test_factors()`.

## **Input data**

Split the example data into a training and a test set. The deconvolution is computed once for all the samples: it is computed independently for each sample, so this does not leak information between samples.

```{r, eval = FALSE}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto

deconv <- multideconv::compute.deconvolution(
  raw.counts,
  methods    = c("Quantiseq", "Epidish"),
  normalized = TRUE,
  return     = FALSE
)

index <- caret::createDataPartition(traitdata$Best.Confirmed.Overall.Response, p = 0.8, list = FALSE)

raw.counts_train <- raw.counts[, index]
traitdata_train  <- traitdata[index, ]
deconv_train     <- deconv[index, ]

raw.counts_test  <- raw.counts[, -index]
traitdata_test   <- traitdata[-index, ]
deconv_test      <- deconv[-index, ]
```

## **Training**

Pass `prepare_celltfusion_folds()` as `fold_construction_fun`, and its arguments through `fold_construction_args_fixed`:

- `deconv` — deconvolution of the training samples, in the same order as `features_train`.
- `raw.counts` — count matrix of the training samples. `pipeML` converts gene names with `make.names()` (e.g. `HLA-A` becomes `HLA.A`); passing the counts keeps the original gene symbols.
- any other `CellTFusion()` argument, e.g. `TF.collection`, `minMod` or `corr`.

```{r, eval = FALSE}
ml_res <- pipeML::compute_features.training.ML(
  features_train = t(raw.counts_train),
  task_type      = "classification",
  target_var     = traitdata_train$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  metric         = "AUROC",
  k_folds        = 5,
  n_rep          = 5,
  ncores         = 2,
  fold_construction_fun        = prepare_celltfusion_folds,
  fold_construction_args_fixed = list(deconv     = deconv_train,
                                      raw.counts = raw.counts_train,
                                      TF.collection = "CollecTRI")
)
```

The final model is trained on the latent factors learned on all the training samples. The corresponding `CellTFusion()` result is returned in `ml_res$Custom_output`.

## **Prediction on an independent cohort**

Project the test samples onto the latent factors of `ml_res$Custom_output`, then predict:

```{r, eval = FALSE}
features_test <- data.frame(project_test_factors(ml_res$Custom_output, deconv_test))

pred <- pipeML::compute_prediction(
  model          = ml_res$Model,
  test_data      = features_test,
  target_var     = traitdata_test$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  task_type      = "classification"
)

pred$AUC$AUROC
```

## **Survival outcomes**

The same function is used for survival models, with the follow-up time and event of each sample:

```{r, eval = FALSE}
surv_res <- pipeML::compute_features.training.ML(
  features_train = t(raw.counts_train),
  task_type      = "survival",
  time_var       = traitdata_train$PFS,
  event_var      = traitdata_train$PFS_event,
  k_folds        = 5,
  n_rep          = 5,
  ncores         = 2,
  fold_construction_fun        = prepare_celltfusion_folds,
  fold_construction_args_fixed = list(deconv = deconv_train, raw.counts = raw.counts_train)
)

features_test <- data.frame(project_test_factors(surv_res$Custom_output, deconv_test))

surv_pred <- pipeML::compute_prediction(
  model     = surv_res$Model,
  test_data = features_test,
  task_type = "survival",
  time_var  = traitdata_test$PFS,
  event_var = traitdata_test$PFS_event
)

surv_pred$c_index
```

## **Multi-cohort data**

For samples from several cohorts, pass `batch = TRUE`, the metadata of the training samples (`coldata`, in the same order as `features_train`) and the cohort column (`batch_id`). `CellTFusion()` then corrects for cohort in each fold (see [Multi-cohort analysis](a6_batch_analysis.html)), and the test samples of each cohort are projected separately, centered on their own means. With `LODO = TRUE`, `pipeML` leaves one cohort out in each fold:

```{r, eval = FALSE}
ml_res <- pipeML::compute_features.training.ML(
  features_train = t(raw.counts_train),
  task_type      = "classification",
  target_var     = coldata_train$Response,
  trait.positive = "R",
  metric         = "AUROC",
  LODO           = TRUE,
  batch_var      = coldata_train$Cohort,
  fold_construction_fun        = prepare_celltfusion_folds,
  fold_construction_args_fixed = list(deconv     = deconv_train,
                                      raw.counts = raw.counts_train,
                                      coldata    = coldata_train,
                                      batch      = TRUE,
                                      batch_id   = "Cohort")
)
```

For a model trained with `batch = TRUE`, project each new cohort separately with `project_test_factors()`, as its samples are centered on their own means.

**Note:** `prepare_celltfusion_folds()` runs `CellTFusion()` in every fold. Use `ncores` in `fold_construction_args_fixed` to compute several folds in parallel (the package must be installed).
