---
title: "Machine learning workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine learning workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

`CellTFusion` latent factors (`res$Latent_spaces$Z`) can be used as features to predict clinical outcomes. This tutorial trains a classifier on a training cohort and applies it to an independent cohort by projecting the new samples onto the latent factors learned in training, with `project_test_factors()`.

Model training uses the [`pipeML`](https://github.com/VeraPancaldiLab/pipeML) package:

```{r, eval = FALSE}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/pipeML")
```

## **Input data**

Split the example data into a training and a test set:

```{r, eval = FALSE}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto

index <- caret::createDataPartition(
  traitdata[, "Best.Confirmed.Overall.Response"],
  times = 1, p = 0.8, list = FALSE
)

traitdata_train  <- traitdata[index, ]
raw.counts_train <- raw.counts[, index]

traitdata_test   <- traitdata[-index, ]
raw.counts_test  <- raw.counts[, -index]
```

## **Training**

Run `CellTFusion()` on the training set only:

```{r, eval = FALSE}
res_train <- CellTFusion(
  raw.counts     = raw.counts_train,
  normalized     = TRUE,
  deconv_methods = c("Quantiseq", "Epidish"),
  cancer_type    = "skcm",
  file_name      = "Train",
  return         = TRUE
)
```

The latent factor scores are the features. `pipeML::compute_features.training.ML()` trains and cross-validates several classifiers and keeps the best one:

```{r, eval = FALSE}
ml_res <- pipeML::compute_features.training.ML(
  features_train = data.frame(res_train$Latent_spaces$Z),
  task_type      = "classification",
  target_var     = traitdata_train$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  metric         = "AUROC",
  k_folds        = 5,
  n_rep          = 10,
  ncores         = 2,
  return         = TRUE
)
```

## **Projection of an independent cohort**

`project_test_factors()` computes the latent factors of new samples from their deconvolution, using the cell groups and factors learned on the training set. Nothing is re-estimated on the test samples, so no information leaks from the test set into the features.

```{r, eval = FALSE}
deconv_test <- multideconv::compute.deconvolution(
  raw.counts_test,
  methods    = c("Quantiseq", "Epidish"),
  normalized = TRUE,
  return     = FALSE
)

features_test <- data.frame(project_test_factors(res_train, deconv_test))
```

**Note:** if the model was trained with `batch = TRUE` (see [Multi-cohort analysis](a6_batch_analysis.html)), the test samples are centered on their own means, as each training cohort was. Project one cohort at a time, with enough samples to estimate its mean.

## **Prediction**

```{r, eval = FALSE}
pred <- pipeML::compute_prediction(
  model          = ml_res$Model,
  test_data      = features_test,
  target_var     = traitdata_test$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  task_type      = "classification"
)

pred$AUC$AUROC
```

## **Cross-validation with features recomputed in each fold**

For an unbiased estimate of performance, the `CellTFusion` features should be learned inside each cross-validation fold: on the training part of the fold, then projected onto its test part. `pipeML` supports this through its `fold_construction_fun` argument. The function below computes the features in both modes and can be used inside such a fold-construction function:

```{r, eval = FALSE}
compute_features_modular <- function(data, structure = NULL, deconv = NULL,
                                     coldata = NULL, cancer_type = "skcm",
                                     TF.collection = "CollecTRI", file_name = NULL,
                                     final_training = FALSE, normalized = TRUE) {
  if (is.null(structure)) {
    # Training: learn cell groups and latent factors on the training samples
    structure <- CellTFusion(
      raw.counts    = t(data),
      deconv        = deconv,
      normalized    = normalized,
      coldata       = coldata,
      cancer_type   = cancer_type,
      TF.collection = TF.collection,
      file_name     = file_name,
      return        = final_training,
      verbose       = FALSE
    )
    features <- data.frame(structure$Latent_spaces$Z)
  } else {
    # Test: project the test samples onto the training structure
    features <- data.frame(project_test_factors(structure, deconv))
  }
  list(features = features, structure = structure)
}
```

See the `pipeML` documentation for the full fold-construction interface.
