---
title: "Getting Started with CellTFusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CellTFusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

## Overview

`CellTFusion` integrates immune cell-type deconvolution with transcription factor (TF)–gene regulatory networks to characterize immune cell states in the tumor microenvironment from bulk RNA-seq data.

Starting from a raw count matrix, the pipeline produces **latent factors** — compact representations of the TME landscape that can be tested for clinical associations, mapped to known cancer meta-programs, and used as features for machine learning.

## Pipeline steps

| Step | Function | Article |
| --- | --- | --- |
| Cell-type deconvolution | `multideconv::compute.deconvolution()` | [Feature Computation](articles/01-feature-computation.html) |
| TF activity inference | `compute.TFs.activity()` | [Feature Computation](articles/01-feature-computation.html) |
| TF module construction | `compute.WTCNA()` | [Feature Computation](articles/01-feature-computation.html) |
| Pathway activity scoring | `compute.pathway.activity()` | [Feature Computation](articles/01-feature-computation.html) |
| Cell group construction | `construct_cell_groups()` | [Cell Groups & Latent Factors](articles/02-cell-groups.html) |
| Latent factor extraction | `compute.latent_factors()` | [Cell Groups & Latent Factors](articles/02-cell-groups.html) |
| Cell niche derivation | `compute_cells_niches()` | [Cell Groups & Latent Factors](articles/02-cell-groups.html) |
| Hallmark GSEA per factor | `compute_factor_gsea()` | [TME State Characterisation](articles/03-tme-states.html) |
| Meta-program mapping | `map_factors_to_metaprograms()` | [TME State Characterisation](articles/03-tme-states.html) |
| TME subtype annotation | `map_factors_to_TME()` | [TME State Characterisation](articles/03-tme-states.html) |
| Clinical association testing | `scores.stat.analysis()` | [Statistical Analysis](articles/04-analysis.html) |
| Survival analysis | `compute.survival.analysis()` | [Statistical Analysis](articles/04-analysis.html) |
| Test-set projection | `project_test_factors()` | [Machine Learning](articles/06-machine-learning.html) |
| ML model training | `pipeML::compute_features.training.ML()` | [Machine Learning](articles/06-machine-learning.html) |
| Multi-cohort (batch) analysis | `batch = TRUE` in `CellTFusion()` | [Batch/Multi-cohort Analysis](articles/07-batch-analysis.html) |

The full pipeline — every step above run in the right order in a single call — is available through the `CellTFusion()` wrapper function; see the package [README](https://VeraPancaldiLab.github.io/CellTFusion/) for usage examples, including unsupervised, supervised, and multi-cohort modes.

## Quick start

Load the pre-packaged example data and run the full pipeline in one call:

```{r}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto
```

```{r, eval = FALSE}
res <- CellTFusion(
  raw.counts    = raw.counts,
  normalized    = TRUE,
  coldata       = traitdata,
  task          = "unsupervised",
  deconv_methods = c("Quantiseq", "Epidish"),
  cancer_type   = "skcm",
  corr          = 0.7,
  pval          = 0.05,
  file_name     = "Tutorial",
  return        = TRUE
)

# Latent factor scores — use for stat tests and ML
head(res$Latent_spaces$Z)

# TME state annotations
head(res$TME_states)
```

Follow the articles linked in the table above for a step-by-step explanation of each stage.

## Installation

```{r, eval = FALSE}
remotes::install_github("VeraPancaldiLab/CellTFusion")
```

## References
