---
title: "CellTFusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellTFusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`CellTFusion` integrates immune cell type deconvolution with transcription factor (TF)–gene regulatory networks to characterize immune cell states in the tumor microenvironment (TME) from bulk RNA-seq data.

Starting from a count matrix, `CellTFusion` builds **cell groups** — sets of deconvolution features whose abundance follows the activity of a TF co-activity module — and summarizes them into **latent factors**, a compact representation of the TME that can be annotated as TME states, tested against clinical variables and used as features for machine learning.

## Installation

To avoid GitHub API rate limit issues, set up a Personal Access Token (PAT) before installing:

```{r, eval = FALSE}
# install.packages(c("usethis", "gitcreds"))
usethis::create_github_token()
gitcreds::gitcreds_set()
```

Install `CellTFusion` from GitHub:

```{r, eval = FALSE}
# install.packages("pak")
pak::pkg_install("VeraPancaldiLab/CellTFusion")
```

## Quick start

The `CellTFusion()` wrapper runs the whole pipeline in one call, using the example data shipped with the package. Intermediate results and plots are saved in a `Results/` folder in the working directory.

```{r, eval = FALSE}
library(CellTFusion)

res <- CellTFusion(
  raw.counts     = CellTFusion::raw.counts.tuto,
  normalized     = TRUE,
  deconv_methods = c("Quantiseq", "Epidish"),
  cancer_type    = "skcm",
  file_name      = "Tutorial"
)

head(res$Latent_spaces$Z)  # latent factor scores (samples x factors)
res$TME_states             # mapping of each latent factor to a TCGA meta-program
```

## Pipeline

| Step | Function | Tutorial |
| --- | --- | --- |
| Cell type deconvolution | `multideconv::compute.deconvolution()` | [Feature computation](a1_feature_computation.html) |
| TF activity inference | `compute.TFs.activity()` | [Feature computation](a1_feature_computation.html) |
| TF co-activity modules | `compute.WTCNA()` | [Feature computation](a1_feature_computation.html) |
| Pathway activity | `compute.pathway.activity()` | [Feature computation](a1_feature_computation.html) |
| Cell groups | `construct_cell_groups()` | [Cell groups and latent factors](a2_cell_groups.html) |
| Latent factors | `compute.latent_factors()` | [Cell groups and latent factors](a2_cell_groups.html) |
| Cell niches | `compute_cells_niches()` | [Cell groups and latent factors](a2_cell_groups.html) |
| Hallmark GSEA per factor | `compute_factor_gsea()` | [TME state characterization](a3_tme_states.html) |
| Meta-program mapping | `map_factors_to_metaprograms()` | [TME state characterization](a3_tme_states.html) |
| TME subtype annotation | `map_factors_to_TME()` | [TME state characterization](a3_tme_states.html) |
| Clinical associations | `scores.stat.analysis()` | [Statistical analysis](a4_statistical_analysis.html) |
| Survival analysis | `compute.survival.analysis()` | [Statistical analysis](a4_statistical_analysis.html) |
| Cross-validation with `pipeML` | `prepare_celltfusion_folds()` | [Machine learning workflows](a5_machine_learning.html) |
| Projection of new cohorts | `project_test_factors()` | [Machine learning workflows](a5_machine_learning.html) |
| Multi-cohort analysis | `CellTFusion(batch = TRUE)` | [Multi-cohort analysis](a6_batch_analysis.html) |

## Tutorials

Step-by-step tutorials are available in the **Articles** section of the navigation bar:

- **[Feature computation](a1_feature_computation.html)** — cell type deconvolution, TF activity, TF co-activity modules and pathway activity
- **[Cell groups and latent factors](a2_cell_groups.html)** — build cell groups, extract latent factors and characterize cell niches
- **[TME state characterization](a3_tme_states.html)** — Hallmark GSEA, TCGA meta-programs and TME subtypes
- **[Statistical analysis](a4_statistical_analysis.html)** — associations with clinical variables and survival
- **[Machine learning workflows](a5_machine_learning.html)** — use latent factors as features and project independent cohorts
- **[Multi-cohort analysis](a6_batch_analysis.html)** — correct for cohort effects with `batch = TRUE`

## Shiny app

`CellTFusion` includes an interactive app to run the pipeline on the example data or on your own data:

```{r, eval = FALSE}
shiny::runApp(system.file("shiny", package = "CellTFusion"))
```

## Citation

If you use `CellTFusion` in a scientific publication, please cite:

> Hurtado, M., & Pancaldi, V. (2026). *CellTFusion: A transcriptional regulatory network framework for the identification of functional multicellular states from bulk RNA-seq data.* bioRxiv. <https://doi.org/10.64898/2026.06.30.735682>
