---
title: "CellTFusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellTFusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial demonstrates how to use the `CellTFusion` package and explains the main functions of the pipeline. We walk through each step of the pipeline, explain key parameters, and illustrate how to interpret and save results. For more information about the methods we invite to read the main paper of the tool.

```{r setup}
library(CellTFusion)
```

Load the pre-packaged example data included in CellTFusion:
```{r}
raw.counts = CellTFusion::raw.counts.tuto ## Corresponds to bulk RNAseq counts
traitdata = CellTFusion::traitdata.tuto ## Corresponds to the clinical data per sample
```

CellTFusion requires two primary inputs: cell-type deconvolution proportions and transcription factor (TF) activity scores. To obtain these features, we will leverage publicly available tools designed for this purpose.

## **Cell-type deconvolution**

To compute cell-type proportions, we will use the `multideconv` R package. This tool integrates multiple deconvolution algorithms and signature matrices to estimate cell-type abundances from bulk RNA-seq data. For more details, visit the [multideconv GitHub repository](https://github.com/VeraPancaldiLab/multideconv).

```{r}
deconv = multideconv::compute.deconvolution(raw.counts, methods = c("Quantiseq", "Epidish", "DeconRNASeq"), normalized = T, return = F)
```

```{r}
head(deconv[,1:5])
```

## **TF activity inference**

To infer transcription factor activity, we will use the `viper` package (@Alvarez2016) in combination with the `CollecTRI` regulon database (@10.1093/nar/gkad841). This approach estimates TF activity based on the expression levels of their downstream target genes.
```{r}
#We first normalize by log2(TPM + 1)
counts.norm = data.frame(ADImpute::NormalizeTPM(raw.counts, log = T)) 
tfs = compute.TFs.activity(counts.norm, TF.collection = "CollecTRI")
```

```{r}
head(tfs[,1:5])
```

Once features have been calculated we can start analyzing our data using CellTFusion

Before constructing the cell groups, we first reduce the TF activity matrix into modules of TFs that exhibit similar activity patterns across samples. For this, we apply a method inspired by the Weighted Gene Correlation Network Analysis (WGCNA) approach (@langfelder2008wgcna), adapted here for TF activity, which we refer to as Weighted TF Correlation Network Analysis (WTCNA).

```{r}
network = compute.WTCNA(tfs, corr_mod = 0.8, clustering.method = "ward.D2", return = T) 
```

To explore how clinical variables are associated with the TF modules, we can run the following analysis. This function will generate and save the corresponding plots in the `Results/` directory:

```{r}
compute.metadata.association(network[[1]], traitdata, pval = 0.05, file.name = "Tutorial", width = 10) 
```

To functionally characterize the TF modules, we provide a utility to estimate pathway activities using a multivariate linear model (MLM) implemented via the `decoupleR` package (@10.1093/bioadv/vbac016). By default, the `PROGENy` pathway database (@Schubert2018) is used; however, users can supply a custom gene set, in which case Gene Set Variation Analysis (GSVA) will be applied instead.

```{r}
pathways = compute.pathway.activity(counts.norm)
```

To explore the relationship between features, users can run the following function. This will generate and save a labeled heatmap in the `Results/` directory, displaying the correlation between the two feature sets along with significance annotations.

```{r}
compute.modules.relationship(network[[1]], pathways, "Pathways_Progeny-TFs_Modules", width = 15)
```

If pathway analysis alone does not provide sufficient insight into the biological relevance of each module, users can perform Over-Representation Analysis (ORA) using the Reactome database. The function below first identifies hub transcription factors (TFs) for each module based on module membership (MM) and node degree thresholds. It then performs enrichment analysis using the target genes of these hub TFs. Resulting dot plots are automatically saved in the `Results/` directory.

```{r, eval = FALSE}
hub_tfs = identify_hub_TFs(t(tfs), network, MM_thresh = 0.8, degree_thresh = 0.9)
compute.modules.enrichment(counts.norm, hub_tfs)
```

Once the TF modules are defined and functionally characterized, we further refine the cell-type deconvolution results to reduce dimensionality and enhance statistical power in downstream association analyses. To achieve this, we use the `compute.deconvolution.analysis()` function from the `multideconv` R package. This function identifies groups of cell types with similar abundance patterns across samples and returns a reduced feature matrix.

```{r}
dt = multideconv::compute.deconvolution.analysis(deconv, corr = 0.7, seed = 123) 
dt = multideconv::deconvolution_dictionary(dt, pathways) ## Apply dictionary of deconvolution (To be added in compute.deconvolution.analysis() soon)
#Association of TFs modules with deconvolution 
compute.modules.relationship(network[[1]], dt[[1]], "Deconvolution-TFs_Modules", vertical = T, height = 30, width = 10, pval = 0.05)
```

## **Construct cell groups**

Cell groups construction is the main part of CellTFusion. Using the previously calculated features, we identify clusters of cells that share similar biological activity patterns, reflected by the TF modules scores across samples.

Supervised analysis

In this approach, we incorporate clinical traits to guide the identification of cell groups associated with specific phenotypes (e.g., treatment response).

```{r}
cell_groups = construct_cell_groups(counts.norm, tfs, deconv, network, dt, traitdata, pval = 0.05, 
                                    trait = "Best.Confirmed.Overall.Response", positive = "CR")
```

Unsupervised analysis

Alternatively, we can identify cell groups solely based on molecular features without using clinical annotations.

```{r}
cell_groups = construct_cell_groups(counts.norm, tfs, deconv, network, dt, pval = 0.05)
```

## **Analize results**

To evaluate the association between cell group scores and a clinical trait, you can perform statistical tests such as Fisher’s exact test, ANOVA, or any other supported method (Wilcoxon, Kruskal–Wallis, t-test) using the unified function `cell.groups.stat.analysis()`.

```{r, eval = FALSE}
res_fisher <- cell.groups.stat.analysis(cell.groups = cell_groups, coldata = traitdata, 
                                        trait = "Best.Confirmed.Overall.Response", method = "fisher", pval = 0.05)

res_anova <- cell.groups.stat.analysis(cell.groups = cell_groups, coldata = traitdata,
                                       trait = "Best.Confirmed.Overall.Response", method = "anova", pval = 0.05)

res_wilcox <- cell.groups.stat.analysis(cell.groups = cell_groups, coldata = traitdata, 
                                        trait = "Best.Confirmed.Overall.Response", method = "wilcox", pval = 0.05)

res_krustal <- cell.groups.stat.analysis(cell.groups = cell_groups, coldata = traitdata,
                                         trait = "Best.Confirmed.Overall.Response", method = "krustal", pval = 0.05)
```

## **One-step CellTFusion**

For convenience, we offer an all-in-one function that automates the entire CellTFusion workflow—calculating features, performing intermediate analyses, and ultimately returning the cell group scores. To run this streamlined pipeline, simply execute:

```{r}
res <- CellTFusion(
  raw.counts = raw.counts,
  normalized = TRUE,
  coldata = traitdata, # Optional metadata
  trait = "Best.Confirmed.Overall.Response",  # Optional for supervised analysis
  trait.positive = "CR",                      # Define positive class for trait
  deconv_methods = c("Quantiseq", "Epidish", "DeconRNASeq"), # Choose from Quantiseq, Epidish, DeconRNASeq, DWLS, CibersortX
  file_name = "TestRun",
  corr = 0.7,
  pval = 0.05,
  high_corr_groups = 0.85,
  return = FALSE
)
```

## **Machine Learning models**

Below are examples demonstrating how the computed cell groups can be leveraged to train machine learning models for predicting specific clinical traits.

We use the R package `pipeML` for model training and prediction. For more details, visit the [pipeML GitHub repository](https://github.com/VeraPancaldiLab/pipeML).

To start, we split our dataset into training and testing sets as follows:
```{r}
index = caret::createDataPartition(traitdata[,"Best.Confirmed.Overall.Response"], times = 1, p = 0.8, list = FALSE) 

# Train cohort
traitData_train = traitdata[index, ]
raw.counts_train = raw.counts[,index]

# Test cohort
traitData_test = traitdata[-index, ]
raw.counts_test = raw.counts[,-index]
```

We performed CellTFusion in the training set

```{r}
res_training <- CellTFusion(
  raw.counts = raw.counts_train,
  normalized = TRUE,
  coldata = traitData_train, # Optional metadata
  trait = "Best.Confirmed.Overall.Response",  # Optional supervised analysis
  trait.positive = "PD",                      # Define positive class for trait
  deconv_methods = c("Quantiseq", "Epidish", "DeconRNASeq"), # Choose from Quantiseq, Epidish, DeconRNASeq, DWLS, CibersortX
  file_name = "TestRun",
  corr = 0.7,
  pval = 0.05,
  high_corr_groups = 0.85,
  return = FALSE
)
```

Next, we train the machine learning model using the cell group scores:

```{r}
library(caret)
library(pipeML)
res = pipeML::compute_features.training.ML(features_train = res_training$Cell_groups[[1]], 
                                           target_var = traitData_train$Best.Confirmed.Overall.Response, 
                                           trait.positive = "PD", 
                                           metric = "AUROC", 
                                           task_type = "classification",
                                           stack = F, k_folds = 2, 
                                           n_rep = 2,  
                                           ncores = 2,
                                           return = F)
```

We replicate the trained cell groups into an independent dataset

```{r}
## Compute deconvolution in the independent set
deconv_test = multideconv::compute.deconvolution(raw.counts_test, methods = c("Quantiseq", "Epidish", "DeconRNASeq"), normalized = T, return = F)

## Replicate cell groups into independent set
testing_set = compute.test.set(res_training$Processed_deconvolution, res_training$Cell_groups, names(res_training$Cell_groups[[2]]), deconv_test)
```

Predicting clinical trait on the test set

```{r}
pred = pipeML::compute_prediction(res, testing_set, traitData_test$Best.Confirmed.Overall.Response, "PD", stack = F)
head(pred$Metrics[,1:5])
pred$AUC$AUROC
```

## **Custom k-fold cross validation with CellTFusion**

In certain use cases it is essential to carefully design how the training and test splits are created to prevent data leakage. This is especially important for feature construction methods that depend on sample-level correlations or deconvolution results. If such features are computed on the full dataset before splitting, information from the test set can inadvertently influence the training set, leading to inflated performance estimates.

For this purpose, `pipeML` allows the user to define a custom fold construction function. This makes it possible to recompute features at every fold while ensuring that test samples are never included in the feature learning process.

For this purpose, `CellTFusion` provides the function `prepare_CellTFusion_folds()`, which is fully compatible with the `pipeML` package for machine learning model training and evaluation.

The `prepare_CellTFusion_folds()` function handles cross-validation by recomputing `CellTFusion` features inside each fold. This ensures that:

- Test samples are never included in the feature learning process.

- Hyperparameter tuning can be performed safely without introducing bias.

- Parallelization (via foreach and doParallel) allows multiple folds to be processed simultaneously, reducing runtime.

When integrated with `pipeML`, the function can be passed as a custom fold construction function (e.g., `fold_construction_fun` = `prepare_CellTFusion_folds`), ensuring that the cross-validation setup mimics real-world application to unseen data.

Here is an example of how to integrate it into the machine learning workflow:
```{r, eval = FALSE}
universe <- decoupleR::get_collectri(organism = 'human', split_complexes = FALSE)
paths <- decoupleR::get_progeny(organism = 'human', top = 500)

res_groups = compute_features.training.ML(t(raw.counts_train), 
                                          traitData_train$Best.Confirmed.Overall.Response, 
                                          trait.positive = "PD",
                                          metric = "AUROC", 
                                          stack = F, 
                                          k_folds = 3, 
                                          n_rep = 5, 
                                          LODO = F, 
                                          file_name = "Test", 
                                          ncores =  2,                                                                                        
                                          return = T,
                                          fold_construction_fun = prepare_CellTFusion_folds, 
                                          fold_construction_args_fixed = list(deconv = deconv, 
                                                                              universe = universe, 
                                                                              paths = paths, 
                                                                              ncores = 2,
                                                                              normalized = TRUE, 
                                                                              coldata = traitData_train, 
                                                                              trait = "Best.Confirmed.Overall.Response", 
                                                                              trait.positive = "PD"),
                                          fold_construction_args_tunable = list(min_targets_size = c(5, 10, 15, 20), 
                                                                                minMod = c(5, 10, 15, 20), 
                                                                                corr_mod = c(0.7, 0.8, 0.9), 
                                                                                corr = c(0.7, 0.8, 0.9), 
                                                                                high_corr_groups = 0.9))
```

In this setup:

* fold_construction_args_fixed defines parameters that are kept constant during the fold construction.

* fold_construction_args_tunable defines the hyperparameters that will be tuned within the cross-validation process.

👉 For more details on what these arguments mean and how to set them up, please refer to the [pipeML](https://verapancaldilab.github.io/pipeML/articles/pipeML.html) tutorial.

# References
