---
title: "Getting Started with CellTFusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CellTFusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: yes
colorlinks: yes
biblio-style: apalike
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

## Overview

`CellTFusion` integrates immune cell-type deconvolution with transcription factor (TF)–gene regulatory networks to characterize immune cell states in the tumor microenvironment from bulk RNA-seq data.

The pipeline takes raw counts and optional clinical metadata as input, and produces **cell group scores** — biologically coherent clusters of samples defined by shared TF activity and cell-type composition patterns.

## Pipeline steps

| Step                    | Function                                        | Article                                                                    |
| ----------------------- | ----------------------------------------------- | -------------------------------------------------------------------------- |
| Cell-type deconvolution | `multideconv::compute.deconvolution()`          | [Feature Computation](articles/01-feature-computation.html)                |
| TF activity inference   | `compute.TFs.activity()`                        | [Feature Computation](articles/01-feature-computation.html)                |
| TF module construction  | `compute.WTCNA()`                               | [Feature Computation](articles/01-feature-computation.html)                |
| Pathway activity scoring| `compute.pathway.activity()`                    | [Feature Computation](articles/01-feature-computation.html)                |
| Cell group construction | `construct_cell_groups()`                       | [Cell Group Construction](articles/02-cell-groups.html)                    |
| Statistical association | `scores.stat.analysis()`                        | [Statistical Analysis](articles/03-analysis.html)                          |
| Full pipeline (one call)| `CellTFusion()`                                 | [One-step Pipeline](articles/04-one-step-pipeline.html)                    |
| ML model training       | `pipeML::compute_features.training.ML()`        | [Machine Learning](articles/05-machine-learning.html)                      |

## Quick start

Load the pre-packaged example data and run the full pipeline in one call:

```{r}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto
```

```{r, eval = FALSE}
res <- CellTFusion(
  raw.counts     = raw.counts,
  normalized     = TRUE,
  coldata        = traitdata,
  trait          = "Best.Confirmed.Overall.Response",
  trait.positive = "CR",
  deconv_methods = c("Quantiseq", "Epidish"),
  file_name      = "TestRun",
  corr           = 0.7,
  pval           = 0.05,
  return         = FALSE
)
```

For a step-by-step walkthrough of each stage, follow the articles listed in the table above.

## Installation

```{r, eval = FALSE}
# Install from GitHub
remotes::install_github("VeraPancaldiLab/CellTFusion")
```

# References
