---
title: "Feature Computation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature Computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

CellTFusion requires two primary inputs: **cell-type deconvolution proportions** and **transcription factor (TF) activity scores**. This article walks through each step manually. If you prefer to run everything in one call, see the [One-step Pipeline](05-one-step-pipeline.html) article.

Load the pre-packaged example data:

```{r}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto
```

## Cell-type deconvolution

Cell-type proportions are estimated with the `multideconv` R package, which integrates multiple algorithms (Quantiseq, Epidish, DeconRNASeq, DWLS, CIBERSORTx). For more details visit the [multideconv GitHub repository](https://github.com/VeraPancaldiLab/multideconv).

```{r}
deconv <- multideconv::compute.deconvolution(
  raw.counts,
  methods    = c("Quantiseq", "Epidish"),
  normalized = TRUE,
  return     = FALSE
)
head(deconv[, 1:5])
```

## TF activity inference

TF activity is inferred using the `viper` package [@Alvarez2016] combined with the `CollecTRI` regulon database [@10.1093/nar/gkad841]. The expression matrix must be log-normalized (log2 TPM+1) before this step.

```{r, eval = FALSE}
# Normalize by log2(TPM + 1)
counts.norm <- data.frame(ADImpute::NormalizeTPM(raw.counts, log = TRUE))

# Using the default CollecTRI regulon
tfs <- compute.TFs.activity(
  RNA.counts       = counts.norm,
  TF.collection    = "CollecTRI",   # or "Dorothea", "ARACNE"
  min_targets_size = 5,
  cores            = 3,
  return           = TRUE
)
head(tfs[, 1:5])
```

You can also supply a pre-built regulon via the `universe` argument:

```{r, eval = FALSE}
universe <- decoupleR::get_collectri(organism = "human", split_complexes = FALSE)
tfs <- compute.TFs.activity(counts.norm, universe = universe)
```

## TF module construction (WTCNA)

Before identifying cell groups, TF activity scores are reduced into modules of co-active TFs using Weighted TF Correlation Network Analysis (WTCNA), adapted from WGCNA [@langfelder2008wgcna].

```{r, eval = FALSE}
network <- compute.WTCNA(
  TFs.matrix       = tfs,
  batch            = FALSE,
  network.type     = "signed",
  clustering.method = "ward.D2",
  minMod           = 15,
  corr_mod         = 0.9,
  return           = TRUE
)
```

For multi-cohort (batch) data, pass `batch = TRUE` and provide a list of per-cohort TF matrices. Consensus WTCNA is then applied automatically.

To visualize associations between TF modules and clinical traits:

```{r, eval = FALSE}
compute.metadata.association(
  tfs.modules = network[[1]],
  coldata     = traitdata,
  pval        = 0.05,
  file.name   = "Tutorial",
  width       = 10
)
```

## Pathway activity scoring

Pathway activities are estimated using a multivariate linear model (MLM) via `decoupleR` [@10.1093/bioadv/vbac016] with PROGENy [@Schubert2018] as the default database. A custom gene set can be provided, in which case GSVA is applied instead.

```{r, eval = FALSE}
pathways <- compute.pathway.activity(
  RNA.tpm   = counts.norm,
  return    = TRUE
)
```

To correlate TF modules with pathway activities:

```{r, eval = FALSE}
compute.modules.relationship(
  network[[1]], pathways,
  "Pathways_Progeny-TFs_Modules",
  width = 15
)
```

## Deconvolution dimensionality reduction

Cell types with correlated abundance profiles are grouped to reduce the feature space and improve statistical power:

```{r, eval = FALSE}
dt <- multideconv::compute.deconvolution.analysis(
  deconvolution = deconv,
  corr          = 0.7,
  seed          = 123,
  return        = FALSE
)
```

## Hub TF enrichment (optional)

Hub TFs per module can be identified based on module membership and network degree, and used for Over-Representation Analysis against the Reactome database:

```{r, eval = FALSE}
hub_tfs <- identify_hub_TFs(t(tfs), network, MM_thresh = 0.8, degree_thresh = 0.9)
compute.modules.enrichment(counts.norm, hub_tfs)
```

## References
