---
title: "Feature Computation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature Computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

CellTFusion requires two primary inputs: **cell-type deconvolution proportions** and **transcription factor (TF) activity scores**. This article walks through how to compute both, followed by TF module construction and pathway activity scoring.

Load the pre-packaged example data:

```{r}
raw.counts  <- CellTFusion::raw.counts.tuto
traitdata   <- CellTFusion::traitdata.tuto
```

## Cell-type deconvolution

To compute cell-type proportions, we use the `multideconv` R package, which integrates multiple deconvolution algorithms and signature matrices to estimate cell-type abundances from bulk RNA-seq data. For more details, visit the [multideconv GitHub repository](https://github.com/VeraPancaldiLab/multideconv).

```{r}
deconv <- multideconv::compute.deconvolution(
  raw.counts,
  methods    = c("Quantiseq", "Epidish"),
  normalized = TRUE,
  return     = FALSE
)
head(deconv[, 1:5])
```

## TF activity inference

To infer transcription factor activity, we use the `viper` package [@Alvarez2016] combined with the `CollecTRI` regulon database [@10.1093/nar/gkad841]. This approach estimates TF activity from the expression levels of downstream target genes.

```{r, eval = FALSE}
# Normalize by log2(TPM + 1)
counts.norm <- data.frame(ADImpute::NormalizeTPM(raw.counts, log = TRUE))

universe <- decoupleR::get_collectri(organism = "human", split_complexes = FALSE)
tfs <- compute.TFs.activity(counts.norm, universe = universe)
head(tfs[, 1:5])
```

## TF module construction (WTCNA)

Before constructing cell groups, we reduce the TF activity matrix into modules of TFs that exhibit similar activity patterns across samples. This is done using a Weighted TF Correlation Network Analysis (WTCNA), adapted from WGCNA [@langfelder2008wgcna].

```{r, eval = FALSE}
network <- compute.WTCNA(tfs, corr_mod = 0.8, clustering.method = "ward.D2", return = TRUE)
```

To explore how clinical variables associate with TF modules, use `compute.metadata.association()`. Plots are saved in the `Results/` directory:

```{r, eval = FALSE}
compute.metadata.association(network[[1]], traitdata, pval = 0.05, file.name = "Tutorial", width = 10)
```

## Pathway activity scoring

To functionally characterize TF modules, we estimate pathway activities using a multivariate linear model (MLM) via `decoupleR` [@10.1093/bioadv/vbac016] with the `PROGENy` database [@Schubert2018] by default. A custom gene set can also be supplied, in which case GSVA is applied instead.

```{r, eval = FALSE}
pathways <- compute.pathway.activity(counts.norm)
```

To visualize the relationship between TF modules and pathways, use `compute.modules.relationship()`. This saves a labeled correlation heatmap to `Results/`:

```{r, eval = FALSE}
compute.modules.relationship(network[[1]], pathways, "Pathways_Progeny-TFs_Modules", width = 15)
```

## Hub TF enrichment analysis

If pathway scoring alone is insufficient, you can perform Over-Representation Analysis (ORA) on the Reactome database using hub TFs. The function identifies hub TFs per module (based on module membership and node degree) and runs enrichment on their target genes:

```{r, eval = FALSE}
hub_tfs <- identify_hub_TFs(t(tfs), network, MM_thresh = 0.8, degree_thresh = 0.9)
compute.modules.enrichment(counts.norm, hub_tfs)
```

## Deconvolution dimensionality reduction

To reduce dimensionality in the deconvolution output and improve statistical power downstream, we group cell types with similar abundance patterns using `multideconv`:

```{r, eval = FALSE}
dt <- multideconv::compute.deconvolution.analysis(deconv, corr = 0.7, seed = 123)
dt <- multideconv::deconvolution_dictionary(dt, pathways)

# Correlate deconvolution groups with TF modules
compute.modules.relationship(
  network[[1]], dt[[1]], "Deconvolution-TFs_Modules",
  vertical = TRUE, height = 30, width = 10, pval = 0.05
)
```

# References
