---
title: "Cell Group Construction and Latent Factors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell Group Construction and Latent Factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

This article covers the core analytical steps of CellTFusion: identifying cell groups from TF modules and deconvolution data, extracting latent factors, and deriving cell niches. These steps require outputs from the [Feature Computation](01-feature-computation.html) article:

- `network` — WTCNA TF modules from `compute.WTCNA()`
- `dt` — reduced deconvolution groups from `multideconv::compute.deconvolution.analysis()`
- `latent_spaces` — NMF latent factors from `compute.latent_factors()`

## Construct cell groups

Cell groups are identified by correlating TF module eigengenes with deconvolution subgroups and cutting the resulting dendrogram. The function takes only two primary inputs — `network` and `dt` — both produced in the feature computation step.

```{r, eval = FALSE}
cell_groups <- construct_cell_groups(
  network           = network,
  dt                = dt,
  batch             = NULL,          # pass batch vector if multi-cohort
  pval              = 0.05,
  clustering.method = "ward.D2",
  n_perm            = 999,
  return_dendrogram = FALSE
)
```

For multi-cohort data, pass the batch vector used in `compute.WTCNA()`:

```{r, eval = FALSE}
batch_vec   <- traitdata[, "Cohort"]
cell_groups <- construct_cell_groups(network, dt, batch = batch_vec, pval = 0.05)
```

### Output structure

`construct_cell_groups()` returns a three-element list:

| Element | Description |
| ------- | ----------- |
| `[[1]]` | Data frame of composite scores (samples × cell groups) |
| `[[2]]` | Named list of cell-type compositions per group |
| `[[3]]` | Named list of TF loading vectors per group |

Results are also written to `Results/Cell.groups.composition.csv` and `Results/Cell.groups.scores.csv`.

## Extract latent factors

Cell group scores are decomposed into NMF latent factors to obtain a compact, non-negative representation of the TME landscape. The `rank` argument controls the number of factors; if `NULL`, it is selected automatically.

```{r, eval = FALSE}
latent_spaces <- compute.latent_factors(
  X         = cell_groups[[1]],
  rank      = NULL,
  seed      = 123,
  file_name = "Tutorial",
  return    = TRUE
)
```

The result is a list whose `$Z` element is a samples × factors matrix — the primary input for downstream statistical tests and machine learning.

```{r, eval = FALSE}
# Access the factor scores matrix (samples x factors)
head(latent_spaces$Z)
```

## Extract cell niches

Cell niches are derived from latent factors by enriching each factor for specific cell-type signals from the deconvolution output. This step characterises each latent factor with a biological cell-type context.

```{r, eval = FALSE}
cells_niches <- compute_cells_niches(
  latent_spaces   = latent_spaces,
  dt              = dt,
  cell.groups     = cell_groups,
  enrich_thresh   = 1.5,
  quantile_cutoff = 0.7,
  return          = TRUE,
  file_name       = "Tutorial"
)
```

## Next steps

With `latent_spaces` in hand, you can:

- **Characterise TME states** via GSEA and meta-program mapping → [TME States](03-tme-states.html)
- **Test clinical associations** of latent factors → [Statistical Analysis](04-analysis.html)
- **Train ML models** on `latent_spaces$Z` → [Machine Learning](06-machine-learning.html)
