---
title: "Statistical Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

Once cell groups have been constructed (see [Cell Group Construction](02-cell-groups.html)), you can evaluate their association with clinical traits using a range of statistical tests.

## Statistical association tests

The unified function `scores.stat.analysis()` accepts:

- Cell group scores from `construct_cell_groups()`
- NMF latent factors from `compute.latent_factors()`
- Any samples × features numeric matrix

Supported methods are `"fisher"`, `"anova"`, `"wilcox"`, `"kruskal"`, and `"ttest"`.

```{r, eval = FALSE}
# Fisher's exact test
res_fisher <- scores.stat.analysis(
  scores  = cell_groups,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "fisher",
  pval    = 0.05
)

# ANOVA
res_anova <- scores.stat.analysis(
  scores  = cell_groups,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "anova",
  pval    = 0.05
)

# Wilcoxon rank-sum test
res_wilcox <- scores.stat.analysis(
  scores  = cell_groups,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "wilcox",
  pval    = 0.05
)

# Kruskal–Wallis test
res_kruskal <- scores.stat.analysis(
  scores  = cell_groups,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "kruskal",
  pval    = 0.05
)
```

## Latent factor analysis

`compute.latent_factors()` decomposes cell group scores into NMF latent factors, which can then be passed directly to `scores.stat.analysis()`:

```{r, eval = FALSE}
nmf <- compute.latent_factors(cell_groups)

res_nmf <- scores.stat.analysis(
  scores  = nmf,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "anova",
  pval    = 0.05
)
```

## Clinical metadata association

To explore associations between TF module scores and all available clinical traits simultaneously, use `compute.metadata.association()`. It applies Pearson correlation for continuous traits and ANOVA for categorical traits, and saves a labeled heatmap and violin plots to `Results/`:

```{r, eval = FALSE}
compute.metadata.association(
  tfs.modules = network[[1]],
  coldata     = traitdata,
  pval        = 0.05,
  file.name   = "Tutorial",
  width       = 10
)
```
