---
title: "TME State Characterisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TME State Characterisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

Once latent factors have been extracted (see [Cell Group Construction](02-cell-groups.html)), each factor can be functionally annotated by linking it to known biological programs. This article covers:

1. **Hallmark GSEA** — associate each latent factor with MSigDB Hallmark gene sets
2. **Meta-program mapping** — compare factors to cancer-type-specific reference programs derived from TCGA
3. **TME subtype annotation** — annotate factors with established TME immune subtypes

These steps require:

- `counts.norm` — log-normalized expression matrix (genes × samples)
- `latent_spaces` — output of `compute.latent_factors()`; the `$Z` element is the samples × factors matrix

---

## Hallmark GSEA per latent factor

`compute_factor_gsea()` fits a multivariate limma model using the latent factor scores as covariates, ranks genes by moderated t-statistic per factor, and runs Hallmark GSEA via `fgsea`. A dot plot of the top enriched pathways is saved per factor.

```{r, eval = FALSE}
gsea_results <- compute_factor_gsea(
  RNA.tpm     = counts.norm,        # genes × samples expression matrix
  features_df = latent_spaces$Z,    # samples × factors (from compute.latent_factors)
  plot_dot    = TRUE,
  top_n       = 10,
  file_name   = "Tutorial",
  width       = 8,
  height      = 10
)
```

The result is a list with two elements:

| Element | Description |
| ------- | ----------- |
| `$DE_results` | Named list of limma DEG tables, one per factor |
| `$GSEA_results` | Named list of fgsea result data frames, one per factor |

---

## Map factors to cancer-type meta-programs

`map_factors_to_metaprograms()` compares each factor's Hallmark NES profile against a reference set of meta-programs derived from TCGA bulk RNA-seq data. The reference covers multiple cancer types and allows cross-study interpretation of the identified TME states.

Supported `cancer_type` values: `"skcm"` (melanoma), `"blca"` (bladder cancer), `"luad"` (lung adenocarcinoma).

```{r, eval = FALSE}
mp_mapping <- map_factors_to_metaprograms(
  gsea_study  = gsea_results,
  cancer_type = "skcm",        # match to your cancer type
  plot        = TRUE,
  file_name   = "Tutorial"
)
```

The result contains:

| Element | Description |
| ------- | ----------- |
| `$factor_mapping` | Data frame mapping each study factor to the closest reference meta-program |
| `$reference` | The reference meta-program NES matrix used for comparison |

---

## Annotate factors with TME immune subtypes

`map_factors_to_TME()` uses TCGA sample-level TME subtype annotations to score each latent factor for enrichment in specific immune environments (e.g., immune-desert, immune-excluded, inflamed).

```{r, eval = FALSE}
tme_annotation <- map_factors_to_TME(
  cancer_name = "skcm",
  Z           = latent_spaces$Z,
  plot        = TRUE,
  file_name   = "Tutorial"
)
```

---

## Derive meta-programs from GSEA (unsupervised)

If you have run CellTFusion across multiple cohorts and want to derive consensus meta-programs from scratch rather than mapping to TCGA references, use `derive_meta_programs()`. This clusters factors by their Hallmark NES profiles to identify recurring biological programs.

```{r, eval = FALSE}
meta_programs <- derive_meta_programs(
  gsea_results = gsea_results,
  k            = NULL,       # number of clusters; NULL for automatic selection
  file_name    = "Tutorial",
  plot         = TRUE
)
```

---

## Putting it all together

In practice, this full annotation sequence is run automatically when using the `CellTFusion()` wrapper. The outputs are accessible directly from the result object:

```{r, eval = FALSE}
res <- CellTFusion(raw.counts = raw.counts, cancer_type = "skcm", ...)

# Access GSEA + meta-program mapping results
res$TME_states           # factor-to-meta-program mapping table
res$Metaprograms_reference  # reference NES matrix

# Access latent factors used as input to GSEA
head(res$Latent_spaces$Z)
```
