---
title: "TME State Characterisation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TME State Characterisation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../references.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

Once latent factors have been extracted (see [Cell Group Construction](02-cell-groups.html)), each factor can be functionally annotated by linking it to known biological programs. This article covers:

1. **Hallmark GSEA** — associate each latent factor with MSigDB Hallmark gene sets
2. **Meta-program mapping** — compare factors to cancer-type-specific reference programs derived from TCGA
3. **TME subtype annotation** — annotate factors with established TME immune subtypes

These steps require:

- `counts.norm` — log-normalized expression matrix (genes × samples)
- `latent_spaces` — output of `compute.latent_factors()`; the `$Z` element is the samples × factors matrix

---

## Hallmark GSEA per latent factor

`compute_factor_gsea()` fits a multivariate `limma` model using the latent factor scores as covariates, ranks genes by moderated t-statistic per factor, and runs a pre-ranked Hallmark gene set enrichment analysis via `fgsea` [@Korotkevich2021] against the MSigDB Hallmark collection [@Liberzon2015]. The Hallmark collection summarizes ~50 curated gene sets representing well-defined, non-redundant biological processes (e.g. EMT, interferon response, hypoxia, inflammatory response), making it a natural, interpretable reference to functionally annotate a data-driven latent factor. A dot plot of the top enriched pathways is saved per factor.

```{r, eval = FALSE}
gsea_results <- compute_factor_gsea(
  RNA.tpm     = counts.norm,        # genes × samples expression matrix
  features_df = latent_spaces$Z,    # samples × factors (from compute.latent_factors)
  plot_dot    = TRUE,
  top_n       = 10,
  file_name   = "Tutorial",
  width       = 8,
  height      = 10
)
```

The result is a list with two elements:

| Element | Description |
| ------- | ----------- |
| `$DE_results` | Named list of limma DEG tables, one per factor |
| `$GSEA_results` | Named list of fgsea result data frames, one per factor |

---

## Map factors to cancer-type meta-programs

### Background: what are meta-programs?

A **meta-program (MP)** is a recurrent module of co-expressed genes reflecting a specific transcriptional cell state (e.g. cell cycle, hypoxia, epithelial-mesenchymal transition, interferon response, stress) that recurs *across* tumors, patients, and even cancer types, rather than being specific to one dataset. This concept was systematically characterized by Gavish et al. [@Gavish2023] ("Hallmarks of transcriptional intratumour heterogeneity across a thousand tumours", *Nature*, 2023), who derived a curated, pan-cancer set of intratumour heterogeneity meta-programs from single-cell RNA-seq data spanning many cancer types. CellTFusion builds cancer-type-specific reference meta-programs from bulk TCGA RNA-seq data using the same underlying logic — recurrent, biologically interpretable transcriptional programs — so that latent factors identified in an independent study can be related back to a well-established vocabulary of TME states instead of being described only in dataset-specific terms.

### How `map_factors_to_metaprograms()` works

1. **Reference construction (offline, per cancer type):** for a given TCGA cohort (e.g. `"skcm"` for melanoma, `"blca"` for bladder cancer), a Hallmark GSEA is run per meta-program to build a reference Hallmarks x meta-programs normalized enrichment score (NES) matrix. This reference is shipped with the package as pre-built `.RData` objects, one per supported cancer type.
2. **Study-side profile:** the Hallmark NES profile of each *study* latent factor, computed in the previous step by `compute_factor_gsea()`, is assembled into a Hallmarks x factors matrix (`build_nes_matrix()`).
3. **Matching:** for each study factor, the function computes, for every reference meta-program, the mean NES of the Hallmark gene sets that define that meta-program. The meta-program with the highest mean NES (and NES > 0) is assigned as the factor's `best_MP`.

This effectively asks: *"among all known recurrent TME/tumour transcriptional programs for this cancer type, which one does this latent factor's Hallmark signature resemble most?"* — turning an abstract NMF factor into a biologically named TME state (e.g. "hypoxia", "interferon response", "stromal/CAF-like").

Supported `cancer_type` values (bundled TCGA reference): `"skcm"` (melanoma), `"blca"` (bladder cancer), `"luad"` (lung adenocarcinoma).

```{r, eval = FALSE}
mp_mapping <- map_factors_to_metaprograms(
  gsea_study  = gsea_results,
  cancer_type = "skcm",        # match to your cancer type
  plot        = TRUE,
  file_name   = "Tutorial"
)
```

The result contains:

| Element | Description |
| ------- | ----------- |
| `$factor_mapping` | Data frame with, per study factor: `factor`, `best_MP` (closest reference meta-program), `best_score` (its mean NES), `all_scores` (mean NES against every reference meta-program) |
| `$reference` | The reference Hallmarks x meta-programs NES matrix used for comparison |

A heatmap of factor-to-meta-program scores is saved to `Results/` when `plot = TRUE`.

---

## Annotate factors with TME immune subtypes

`map_factors_to_TME()` uses TCGA sample-level annotations from Bagaev et al. [@Bagaev2021], who defined four conserved, pan-cancer **Molecular Functional Portrait (MFP)** subtypes of the tumor microenvironment from bulk transcriptomes:

- **IE** (Immune Enriched, non-fibrotic) — high immune infiltration, low stromal/fibrotic signal.
- **IE/F** (Immune Enriched, Fibrotic) — high immune infiltration co-occurring with a fibrotic/stromal signature.
- **F** (Fibrotic) — stroma/fibrosis-dominated, low immune infiltration.
- **D** (Depleted) — low immune infiltration and low fibrosis ("cold" tumors).

These subtypes were shown to be associated with differential response to immune checkpoint blockade, making them a clinically relevant vocabulary to annotate CellTFusion latent factors against. For each cancer type, `map_factors_to_TME()` matches TCGA patients to their MFP subtype, then computes, per latent factor, the Spearman correlation between factor scores and each subtype (`cor_IE`, `cor_IEF`, `cor_F`, `cor_D`); the subtype with the strongest positive correlation is assigned as the factor's `best_MFP`.

```{r, eval = FALSE}
tme_annotation <- map_factors_to_TME(
  cancer_name = "skcm",
  Z           = latent_spaces$Z,
  plot        = TRUE,
  file_name   = "Tutorial"
)
```

---

## Derive meta-programs from GSEA (unsupervised)

If you have run CellTFusion across multiple cohorts and want to derive consensus meta-programs from scratch rather than mapping to TCGA references, use `derive_meta_programs()`. This clusters factors by their Hallmark NES profiles to identify recurring biological programs.

```{r, eval = FALSE}
meta_programs <- derive_meta_programs(
  gsea_results = gsea_results,
  k            = NULL,       # number of clusters; NULL for automatic selection
  file_name    = "Tutorial",
  plot         = TRUE
)
```

---

## Putting it all together

In practice, this full annotation sequence is run automatically when using the `CellTFusion()` wrapper. The outputs are accessible directly from the result object:

```{r, eval = FALSE}
res <- CellTFusion(raw.counts = raw.counts, cancer_type = "skcm", ...)

# Access GSEA + meta-program mapping results
res$TME_states           # factor-to-meta-program mapping table
res$Metaprograms_reference  # reference NES matrix

# Access latent factors used as input to GSEA
head(res$Latent_spaces$Z)
```

## References
