---
title: "Statistical Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

This article demonstrates how to test associations between CellTFusion latent factors and clinical variables. The primary input is `res$Latent_spaces` — the NMF latent factor object returned by `compute.latent_factors()` or the `CellTFusion()` wrapper.

Load example data:

```{r}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto
```

## Association between latent factors and clinical traits

`scores.stat.analysis()` tests the association between latent factor scores and a clinical variable. It accepts the full `latent_spaces` object (not just `$Z`) as its `scores` argument.

Supported `method` values: `"fisher"`, `"wilcox"`, `"anova"`, `"kruskal"`, `"ttest"`.

```{r, eval = FALSE}
# Run CellTFusion to obtain latent spaces
res <- CellTFusion(
  raw.counts  = raw.counts,
  normalized  = TRUE,
  cancer_type = "skcm",
  return      = TRUE
)

# Wilcoxon rank-sum test (binary response variable)
sig_factors <- scores.stat.analysis(
  scores  = res$Latent_spaces,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "wilcox",
  pval    = 0.05
)

# Fisher's exact test
sig_factors_fisher <- scores.stat.analysis(
  scores  = res$Latent_spaces,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "fisher",
  pval    = 0.05
)

# ANOVA (multi-level categorical variable)
sig_factors_anova <- scores.stat.analysis(
  scores  = res$Latent_spaces,
  coldata = traitdata,
  trait   = "Best.Confirmed.Overall.Response",
  method  = "anova",
  pval    = 0.05
)
```

The function returns a list of significant factor results along with visualisations (box plots, violin plots) saved to `Results/`.

## Direct access to factor scores

If you want to work with the factor scores directly (e.g., for custom plots or downstream modelling), access the `$Z` matrix:

```{r, eval = FALSE}
# samples x factors matrix
factor_scores <- data.frame(res$Latent_spaces$Z)
head(factor_scores)
```

## TF module — clinical trait association

To visualise associations between TF module eigengenes (rather than latent factors) and all available clinical traits simultaneously, use `compute.metadata.association()`. It runs Pearson correlation for continuous traits and ANOVA for categorical traits, and saves a labelled heatmap and violin plots to `Results/`:

```{r, eval = FALSE}
compute.metadata.association(
  tfs.modules = res$TF_network[[1]],
  coldata     = traitdata,
  pval        = 0.05,
  file.name   = "Tutorial",
  width       = 10
)
```

## TF modules — pathway correlation

To explore the relationship between TF module scores and pathway activities:

```{r, eval = FALSE}
compute.modules.relationship(
  matA      = res$TF_network[[1]],
  matB      = res$Pathways_scores,
  file_name = "Pathways_vs_TF_modules",
  width     = 15
)
```

## Survival analysis

`compute.survival.analysis()` tests whether Kaplan-Meier survival curves differ significantly between groups, using the `survival` and `survminer` packages. It supports two modes:

1. **Predefined groups** (`group_column`) — e.g. a clinical risk group, treatment arm, or a cluster/cell-group assignment already present in `coldata`.
2. **Automatic feature screening** (`features`) — e.g. `res$Latent_spaces$Z`, TF module scores, or cell group scores. Each feature/column is split into High/Low groups by a quantile cutoff (`thres`), and only features whose log-rank test is significant (`p.value`) are returned.

Both modes require `traitdata` to include a time-to-event column (`PFS`) and an event indicator column (`PFS_event`, 1 = event, 0 = censored).

```{r, eval = FALSE}
# Mode 1: compare survival between predefined clinical groups
surv_group <- compute.survival.analysis(
  survival.data = traitdata,
  PFS           = "PFS",
  PFS_event     = "PFS_event",
  group_column  = "Best.Confirmed.Overall.Response",
  file_name     = "Tutorial"
)

surv_group$p_value        # log-rank test p-value
surv_group$median_PFS      # median survival per group
surv_group$km_plot          # survminer::ggsurvplot object
```

```{r, eval = FALSE}
# Mode 2: screen latent factors for significant survival splits
surv_factors <- compute.survival.analysis(
  survival.data = traitdata,
  PFS           = "PFS",
  PFS_event     = "PFS_event",
  features      = data.frame(res$Latent_spaces$Z),
  p.value       = 0.05,
  thres         = 0.5,        # median split; use e.g. 0.75 for a top-quartile-vs-rest split
  file_name     = "Tutorial"
)
```

For every significant result, a Kaplan-Meier plot with a risk table is saved to `Results/SurvPlot_<group-or-feature>_<file_name>.svg`.

This is the same function used for survival analysis in the CellTFusion companion study (`CellTFusion_paper`), e.g. to test whether NMF latent factor scores stratify patients by overall/progression-free survival.
