---
title: "One-step Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{One-step Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CellTFusion)
```

For convenience, `CellTFusion()` is an all-in-one wrapper that automates the entire pipeline — computing features, running intermediate analyses, and returning cell group scores — in a single call.

The individual steps (feature computation, cell group construction, statistical analysis) are described in their respective articles:

- [Feature Computation](01-feature-computation.html)
- [Cell Group Construction](02-cell-groups.html)
- [Statistical Analysis](03-analysis.html)

## Running the full pipeline

```{r, eval = FALSE}
library(CellTFusion)

raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto

res <- CellTFusion(
  raw.counts     = raw.counts,
  normalized     = TRUE,
  coldata        = traitdata,
  trait          = "Best.Confirmed.Overall.Response",
  trait.positive = "CR",
  deconv_methods = c("Quantiseq", "Epidish"),
  file_name      = "TestRun",
  corr           = 0.7,
  pval           = 0.05,
  high_corr_groups = 0.85,
  return         = FALSE
)
```

## Key parameters

| Parameter | Description |
|---|---|
| `raw.counts` | Raw or normalized count matrix (genes × samples) |
| `normalized` | Set `TRUE` if counts are already log-normalized |
| `coldata` | Data frame of clinical metadata (optional) |
| `trait` | Column in `coldata` to use for supervised analysis |
| `trait.positive` | Positive class label for the trait |
| `deconv_methods` | Deconvolution algorithms to use |
| `corr` | Correlation threshold for grouping cell types |
| `pval` | P-value threshold for significance filters |
| `high_corr_groups` | Threshold for merging highly correlated cell groups |
| `return` | If `TRUE`, returns the result object instead of writing to disk |

## Output

`CellTFusion()` returns a named list containing:

| Element | Description |
|---|---|
| `$Cell_groups` | Cell group scores and composition |
| `$Processed_deconvolution` | Reduced deconvolution feature matrix |
| `$TF_modules` | WTCNA network output |
| `$Pathways` | PROGENy pathway activity scores |
