---
title: "Machine Learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(CellTFusion)
```

CellTFusion latent factors (`res$Latent_spaces$Z`) serve as features for machine learning models. This article shows how to train a classifier on a training cohort and project it onto an independent test cohort using `project_test_factors()`.

> **Installation:** this article uses the [`pipeML`](https://github.com/VeraPancaldiLab/pipeML) package for model training. Install it from GitHub before running these examples:
> ```r
> remotes::install_github("VeraPancaldiLab/pipeML")
> ```

---

## Train/test split

```{r}
raw.counts <- CellTFusion::raw.counts.tuto
traitdata  <- CellTFusion::traitdata.tuto

index <- caret::createDataPartition(
  traitdata[, "Best.Confirmed.Overall.Response"],
  times = 1, p = 0.8, list = FALSE
)

traitData_train  <- traitdata[index, ]
raw.counts_train <- raw.counts[, index]

traitData_test   <- traitdata[-index, ]
raw.counts_test  <- raw.counts[, -index]
```

## Run CellTFusion on the training set

```{r, eval = FALSE}
res_train <- CellTFusion(
  raw.counts    = raw.counts_train,
  normalized    = TRUE,
  coldata       = traitData_train,
  task          = "unsupervised",
  deconv_methods = c("Quantiseq", "Epidish"),
  cancer_type   = "skcm",
  corr          = 0.7,
  pval          = 0.05,
  file_name     = "Train",
  return        = TRUE
)

# Latent factor scores are in res_train$Latent_spaces$Z
head(res_train$Latent_spaces$Z)
```

## Train a machine learning model

The latent factor scores (`$Z`) are used as features. `pipeML::compute_features.training.ML()` trains and cross-validates classifiers (random forest, elastic net, etc.) and returns AUROC results.

```{r, eval = FALSE}
library(pipeML)

ml_res <- pipeML::compute_features.training.ML(
  features_train = res_train$Latent_spaces$Z,
  target_var     = traitData_train$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  metric         = "AUROC",
  task_type      = "classification",
  stack          = FALSE,
  k_folds        = 5,
  n_rep          = 10,
  ncores         = 2,
  return         = TRUE
)
```

## Project to an independent test set

`project_test_factors()` takes the full training result object (`res_train`) and the deconvolution of the test cohort, and projects the test samples into the training latent space. This ensures no data leakage from the test set into the feature construction.

```{r, eval = FALSE}
# Compute deconvolution on the test cohort
deconv_test <- multideconv::compute.deconvolution(
  raw.counts_test,
  methods    = c("Quantiseq", "Epidish"),
  normalized = TRUE,
  return     = FALSE
)

# Project test samples into training latent space
features_test <- data.frame(project_test_factors(res_train, deconv_test))
head(features_test)
```

## Predict on the test set

```{r, eval = FALSE}
pred <- pipeML::compute_prediction(
  ml_res,
  features_test,
  traitData_test$Best.Confirmed.Overall.Response,
  trait.positive = "PD",
  stack          = FALSE
)

head(pred$Metrics[, 1:5])
pred$AUC$AUROC
```

---

## Custom k-fold cross-validation

For rigorous evaluation, CellTFusion features should be recomputed inside each fold to prevent data leakage. The helper `compute_features_modular()` pattern from the paper scripts wraps `CellTFusion()` + `project_test_factors()` for use inside `pipeML`'s custom fold interface:

```{r, eval = FALSE}
compute_features_modular <- function(data, structure = NULL, deconv = NULL,
                                     coldata = NULL, universe = NULL, paths = NULL,
                                     normalized = TRUE, task = "unsupervised",
                                     cancer_type = "skcm", TF.collection = "CollecTRI",
                                     file_name = NULL, final_training = FALSE,
                                     min_targets_size = 5, minMod = 10,
                                     corr_mod = 0.7, corr = 0.7) {
  if (is.null(structure)) {
    # Training mode: run full CellTFusion
    structure <- CellTFusion(
      raw.counts       = t(data),
      deconv           = deconv,
      normalized       = normalized,
      coldata          = coldata,
      universe         = universe,
      paths            = paths,
      task             = task,
      cancer_type      = cancer_type,
      TF.collection    = TF.collection,
      min_targets_size = min_targets_size,
      minMod           = minMod,
      corr_mod         = corr_mod,
      corr             = corr,
      file_name        = file_name,
      return           = final_training,
      verbose          = FALSE
    )
    features <- data.frame(structure$Latent_spaces$Z)
  } else {
    # Projection mode: map test samples into training latent space
    features <- data.frame(project_test_factors(structure, deconv))
  }
  list(features = features, structure = structure)
}
```

This pattern ensures the TF module structure is learned only from training samples and consistently applied to test samples at each fold.
